---
title: "ica-joining"
format: html
---

```{r}
students_1 <- data.frame(
  student = c("A", "B", "C"),
  class = c("STAT 101", "GEOL 101", "ANTH 101")
)

# Check it out
students_1
```
```{r}
enrollments_1 <- data.frame(
  class = c("STAT 101", "ART 101", "GEOL 101"),
  enrollment = c(18, 17, 24)
)

# Check it out
enrollments_1
```
First, consider the following:

What variable or key do these datasets have in common? Thus by what information can we match the observations in these datasets? class

Relative to this key, what info does students_1 have that enrollments_1 doesn’t? students in the class

Relative to this key, what info does enrollments_1 have that students_1 doesn’t? enrollment in that class

First, try the left_join() function:

```{r}
library(tidyverse)
enrollments_1 |> 
  left_join(students_1)
```

What did this do? What are the roles of students_1 (the left table) and enrollments_1 (the right table)? connected by the left side

What, if anything, would change if we reversed the order of the data tables? Think about it, then try. students and enrollment would switch

```{r}
enrollments_1 |> 
  inner_join(students_1)
```
What did this do? What are the roles of students_1 (the left table) and enrollments_1 (the right table)? combined the two sets, removed student c

What, if anything, would change if we reversed the order of the data tables? Think about it, then try.

```{r}
enrollments_1 |> 
  full_join(students_1)
```
```{r}
enrollments_1 |> 
  semi_join(students_1)
```
What did this do? What info would it give us? removed enrollments and just kept students and class

How does semi_join() differ from inner_join()? inner keeps enrollment

What, if anything, would change if we reversed the order of the data tables? Think about it, then try. only keep enrollment and class

```{r}
enrollments_1 |> 
  anti_join(students_1)
```
What did this do? What info would it give us? the observation with no match on the right side

What, if anything, would change if we reversed the order of the data tables? Think about it, then try. give art 101 


